---
title: "LSIRM 2018 - Introduction to R"
author: "Simon Kiss"
date: '2018-06-19'
output: html_document: 
    toc: true
    toc_float: true
    code_folding: hide

---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning=F, message=F,  results='hide', fig.width=4, fig.height=3, fig.align='center')
```

# Working With Data
## Reading Data In
R has a number of functions meant to read data in from a variety of formats. The core functions are `read.table()` for plain text files, `read.csv()` for .csv files. More elaborate files (e.g. SPSS or Stata) require more advanced functions. The three main libraries that offer functions that read SPSS or Stata files are the foreign, haven, and rio libraries. Foreign is the oldest, haven and rio are newer.  

My recommendation is to use the `rio` library to read SPSS and Stata files in. 


## Reading in our Data set
```{r read-in}
#ces<-rio::import('https://github.com/sjkiss/LSIRM_2018/blob/master/data/ces_data.sav?raw=true')
library(haven)
library(labelled)
ces<-read_sav('data/ces15_data.sav')

```

# Summarizing Data Frames
After reading data in, it is usually a good idea to inspect the data to get a feel for what you have read in. 

There are a few commands that can be useful to inspect data frames.  

```{r, examine-data}
#Different ways of summarizing the data set
#Check the first few rows
head(ces)
#The last few rows
tail(ces)
#Check the structure of the object
str(ces)
#Summarize the data set
summary(ces)
#See the variables names
names(ces)


```

#### Assignment 1 
> 1. How many missing values are there in each variable? Are there any variables that might cause problems in this regard? 
> 2. Are there any values that look suspicious? 

# Looking at value labels.

When you import SPSS or STATA data sets via the data with the `rio` package, R imports valuable metadata associated with each variable, including the variable label and the value labels. The variable and value labels are stored attached to each variable.  This is a litle inconvenient to look through them, so we use a function `gather_attrs()` to move them to the `data.frame()` level. It makes it easier to work with them. 



Then, to get the variable and value labels, we use the following commands: `var_label` for variable labels and `val_labels` for value labels. 
```{r get-labels}
var_label(ces)
val_labels(ces)

```

#### Assignment 2
> 1. Save the variable labels and value labels as two separate objects (`vars` and `vals`) so we can look through them later. 

```{r assignment-2-solution}
vars<-var_label(ces)
vals<-val_labels(ces)
```

#### Assignment 3
> 1. Look at the value labels and take note of what are missing values that need to be Recoded. 

# Recoding variables

```{r assignment-3-solution}
vals
```

It looks like for every variable, except for `p_selfplace`, we can recode values 998 to 1000 to be missing. But in addition, some variables respondents also picked `don't know` or `refuse`. Some might try to make use of those data, but we'll just dump them for the moment. 

The easiest function to recode variables is the `Recode` command in the `car` library. 

Please note: there is a conflict between two recode commands; namely, there exists a `recode` in the `car` library (specified as `car::recode`) and in the `dplyr` library (part of the `tidyverse`, `dplyr::recode`). If one package is loaded *after* the other, then the first package's command is masked.  The best way to avoid this is to use capital-R `Recode` from the `car` library.  That will *always* and *only* call the `car::recode` command.  


```{r, eval=F, echo=T}
Recode(x, "old value=new value")
```

So, for the income inquality variable `p_incineq`, we could safely recode any number larger than 3 to be missing. Or we could specify each individual number. Both of the following commands are equivalent. 

```{r results='markup', echo=T}
#load the car library
library(car)
#Recode 1
test<-car::Recode(as.numeric(ces$p_incineq),"3:1000=NA", as.numeric=T)
#Recode 2
test2<-Recode(as.numeric(ces$p_incineq), "3=NA; 4=NA; 998=NA; 1000=NA", as.numeric=T)
#Summary test
summary(test)
#summary test 2
summary(test2)

```
There are a few things to notice about the `Recode()` command. Overall, the recodes are put in **double-quotes** . Each individual recode is separated with a semi-colon. And normally, if you are recoding character strings, you use single-quotes. But, because NA is a reserved character string denoting missing data, it is unquoted. 

Notice now that, after the recode, the maximum value is 2. We've set missing values to be `NA` and they are not counted when we use `summary()`

We also just dumped that recode result in a quick temporary variable called `test`.  What we really want is to save it in a new variable caled `inequal`. When you execute this command, watch what happens to the environment in the top right corner. 

```{r, results='hide', echo=T}
ces$inequal<-Recode(as.numeric(ces$p_incineq), "3:1000=NA")

```

We can also recode numbers to factors.  In this case, it might make more sense to collapse the `sex_r` variable to `gender` as a categorical variable. 
```{r sex-gender}
#recode sex into gender
ces$gender<-Recode(ces$sex_r, "1='M' ; 2='F'", levels=c('M', 'F'), as.factor=T)

```
There are a few things to note here. First, note that, inside the recodes, numbers are not quoted, but character strings are, with single-quotes. Second, there is an additional argument, `levels` that takes a character vector the order of which is important. The first level of the factor is the reference categor.  Lastly, there is the argument `as.factor` which is logical (i.e. either `TRUE` or `FALSE`). If TRUE, the new variable is stored a s a factor, if FALSE it is stored as just a character string. 



#### Assignment 4
> 1. Recode `income_full`, `p_selfplace` and `education` such that missing values are converted to NA. Save `income_full` as `income` and `p_selfplace` as `ideology`. You will have to go back and forth to `vals` to remind yourself of what the missing values are. Recode education into a new variable, `degree` which has the levels `no degree` and `degree`. Check `vals` to get the range of values right.  

```{r assignment-4-solution}
#education
vals$education
#education missing values
ces$education<-Recode(as.numeric(ces$education), "12:1000=NA")
#degree
#check vals
vals
#Recode eduation to degree
ces$degree<-car::Recode(ces$education, "1:7='No degree' ; 8:11='Degree'", levels=c('no degree', 'degree'), as.factor=T)

#Income missing values
vals$income
ces$income<-Recode(as.numeric(ces$income_full), "6:1000=NA")
#Recode p_selfplace
ces$ideology<-Recode(as.numeric(ces$p_selfplace), "1000=NA")
str(ces)

```


Note: There are ways to automate this and we'll deal with those tomorrow. 

```{r summary-ces}
summary(ces)
```
 

Often, when you do Recodes, it is also useful to change the names of the variables, to something more user friendly. This has the added advantage of also keeping the original variables intact, just incase you make a mistake.


# Describing results
Produce a table of descriptive results.


```{r results='markup'}
library(psych)
desc<-describe(ces)
```

# Plotting
```{r load-tidyverse}
library(tidyverse)

library(ggplot2)
library(dplyr)
```

This is where the fun begins.  

Base R's graphics are built around the `plot()` command.  The tidyverse's `ggplot2` plotting function is much more systematic and built on consistent principles.  It's basically better.

The `ggplot()` command is the basic command which contains the basic parameters of a graph and then specifics are added to it.  Below, we specify the underlying data -- `ces` -- and the two most important aesthetics of a graph, the x and the y variable.  Note: in ggplot2 ``aesthetics'' has a very particular meaning. Namely, it maps particular visual features of a plot (i.e. color, shape of points, type of line, the fill of bars) to the values of variables (see the figure below).

```{r, make-plot, fig.cap='No plot!'}


ggplot(ces, aes(x=income))

```

But nothing appears.  This is because the second core aspect of the plot is the `geom`  These are the graphical types of representing the relationship between x and y (see the next figure).

```{r income-bar,fig.cap='Histogram', fig.align='center'}
#Add barplot

  ggplot(ces, aes(x=income))+
  geom_bar()

```
## Types of geoms()
There are *a ton* of geoms in the `ggplot2` library. See the [cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) for details. 

#### Assignment 6
> 1. Produce a density plot of ideology. Save it as `plot1`. 

```{r assignment-6-solution}

plot1<-  ggplot(ces, aes(x=ideology))+geom_density()

```

You can modify the colors and visual aspects of the plot in the `geom()` section. When you just want to force the plot to show colors and not be linked or mapped to another variable, you just set the aesthetic (e.g. `color`, `linetype`, `size` or `width` to be a particular value). Handily, to update a plot, you can call the plot and just add a replacement `geom()` to it.

```{r show-aesthetics}
plot1+geom_density(col='red')
plot1+geom_density(size=3)
```

Note: not all `geoms()` accept all aesthetics. 
This: 

```{r linetype-aes, fig.cap='Changing linetype on a densit plot does not do anything.'}
plot1+geom_density(linetype=3)
```

does not change anything. 


Making bivariate plots is just a matter of adding in a y-variable for the y-axis.  

```{r, make-bivariate-plot, fig.cap='A basic scatterplot'}


  ggplot(ces, aes(x=income, y=ideology))+
  geom_point()
```

#### Assignment 7
> 1. Make a scatterplot with education on the x-axis and ideology on the y-axis
> 2. Make a boxplot with education on the x-axis and ideology on the y-axis

```{r assignment-solution-7}
#Scatterplot
plot2<-ggplot(ces, aes(x=education, y=ideology))+geom_point()
plot2
#Boxplot
plot3<-ggplot(ces, aes(x=as.factor(education) ,y=ideology))+geom_boxplot()
plot3
```


Very quickly, we can also add in third grouping variables.  Like education and / or gender.

We can do gender as follows.  
The key is to use the `aes()` argument in the `geom()` to map a specific visuable attribute of the plot to values of the grouping variable. 
```{r, plot-4,include=T}
#Make plot 4, save as plot4
plot4<-ggplot(ces, aes(x=income, y=ideology, group=gender))+geom_point(aes(col=gender))
#Print plot4
plot4

```

#### Assignment 8  

1. Instead of modifying the point colors by category of gender, change them by shape. Save it as `plot4a`

```{r assignment-8-solution}
plot4a<-ggplot(ces, aes(x=income, y=ideology, group=gender))+geom_point(aes(shape=gender))
plot4a
```

### Jittering and Smoothing
You can add multiple geoms to the same plot. This is adding `layers`. It wouldn't really make sense to add a bar plot layer to a scatterplot, but here we can `jitter` the points or we can add a `smooth` to beter show the relationships.

```{r plot4b}
plot4b<-plot4+geom_jitter(aes(col=gender))

plot4b
```

The `smooth` geom plots a linear function through points.  

```{r, plot4b-smooth}
plot4b<-plot4+geom_smooth(method='lm', aes(col=gender))
plot4b
```



#### Assignment 9
> 1.. Produce another scatterplot income and ideology but group the points by education using color to differentiate

```{r assignment-9-solution}
plot5<-ggplot(ces, aes(x=income, y=ideology))+geom_point(aes(col=degree))+geom_smooth(aes(col=degree), method='lm')
plot5
```


The amazing thing about R in general and what has been a major driver in it s success is that _every single part_ of the plot is modifiable. Try doing that in Excel or SPSS.  For example, we can change the axis labels and main title with the command `labs`().

```{r}
plot5+labs(x='Income', y='Ideology', title='Ideology on Income, CES 2015')
```

# Bivariate Analysis - Continuous
## Correlations
Correlations are done with the `cor()` command
Here we correlate personal financial situation with views of whether the federal government's impact on your own situation. Note the argument `use` provides different options for dealing with missing values. Here we'll only use complete cases.
```{r correlation}
names(ces)
str(ces)
cor(ces$ideology, ces$income, use='complete.obs')

```

# Bivariate Analysis - Categorical
## Tables

To produce contingency tables, you just use the `table` command. 
Here we check wither men and women differ in their views of whether inequality is a big problem. 


```{r, table-1, results='markup'}
table(ces$gender, ces$inequal)

```

#### Assignment 10
1. Recode`ces$inequal` to be a factor. Check `vals`. Save it as `inequal_cat`. Print the cross tab again

```{r assignment-10-solution}

ces$inequal
#Check values
vals
ces$inequal_cat<-Recode(ces$inequal, "1='Yes' ; 2='No'", as.factor=T)
#Crosstab
tab2<-table(ces$gender, ces$inequal_cat)
```

We can do a lot with this. Two common things to do with this are to produce a chi-squared test and to do a table of percentages.
```{r chi-sq-prop-table, results='markup'}

#Chi-sq.test
chisq.test(tab2)
#For prop.table you can produce either row (1) or column (2) percentgates
#Compare
prop.table(tab2, 1)
#With
prop.table(tab2, 2)
```
```{r}
library(cowplot)
plot_grid(plot1, plot2, plot3, plot4)
```
# Linear Models

The linear model is very easy to fit.  It uses the formula notation which involves a `~`. The dependent variable is on the left, the covariates (i.e. predictors, independent variables are on the right. )

```{r lm-model1, results='markup'}
#Fit linear model
mod1<-lm(ideology ~ income, data=ces)
```
You can add a second term to the model with the `+` and interactions with a `:`.

```{r mod2, results='markup'}
#Model 2
mod2<-lm(ideology~income+education, data=ces)
#summarize
summary(mod2)
```
```{r mod3, results='markup'}
#Model 3
mod3<-lm(ideology~income+education+income:education, data=ces)
#Summarize
summary(mod3)
```

There is a convenience function, `update()`, that allows you to modify an existing formula. 

The first argument to `update()` is a model object. The second is the new model formula. You use `.` to represent the model fitting in the old model and then add whatever you wish. 

```{r, update-mod2}
#New model
mod3a<-update(mod2, .~.+income:education, data=ces)
#Compare
summary(mod3)
summary(mod3a)
```


#### Assignment 11
> 1. Model support for the belief that more or less should be done to reduce income inequality with a series of linear models (ces$p_gap). Check `vals` to remind yourself what the values are. Fit four different models with one variable each: income, ideology, education (as continuous) and gender. Then fit one last model with all four in one model. Then, add one more model where gender is interacted with all three of the other covariates.  You should have six separate models. Summarize each. 

```{r assignment-11-solution}
m1<-lm(p_gap~income, data=ces)
m2<-update(m1, .~+ideology, data=ces)
m3<-update(m2, .~+education, data=ces)
m4<-update(m3, .~+gender, data=ces)
m5<-update(m1, .~.+ideology+education+gender, data=ces)
m6<-update(m5, .~.+gender:income+gender:ideology+gender:education, data=ces)
summary(m1)
summary(m2)
summary(m3)
summary(m4)
summary(m5)
summary(m6)
```

### Reporting Regressions
The `stargazer` library is by far the best package for producing well-formatted regression tables. 


```{r report-regressions, results='asis'}
#load library
library(stargazer)
stargazer(m1, m2, m3, m4, m5, m6, type='text')
```

Crucially, they can be exported as an html file to your desktop and you can insert these easily into Microsoft Word documents. 

```{r export-regressions, results='hide'}
stargazer(m1, m2, m3, m4, m5, m6, type='html', out='model_results.html')
```
There is one significant interaction, `gender:ideology`, so let's fit a new model, where we drop all interactions except that one.  
#### Assignment 12
>1. Fit a model that includes p_gap as the dependent variable, income, ideology education and gender as covariates and an interaction between gender and ideology. Call it m7. Note you can't use `update()` to subract terms.

```{r assignment-12-solution, results='asis'}
#update m5
m7<-update(m5, .~.+ideology:gender, data=ces)
summary(m7)
```
## Regression Diagnostics


## Predicted Values
There are lots of ways to generate predicted values. One of the easiest ways is the `ggeffects` package (see [here](https://cran.r-project.org/web/packages/ggeffects/vignettes/marginaleffects.html) for an introduction). 

```{r}

library(ggeffects)
```

There are two key commands.  `GGaverage` generates average predicted values, `ggpredict` generates marginal predicted values. Both take a model as its first argument and then a character vector of terms.  The first element in the character vector is the term of interest, the second and third values are grouping variables.

Conveniently the output of both are ready for plotting with `ggplot2`.

```{r predicted-values, results='markup'}
m7.preds<-ggpredict(m7, terms=c('ideology [0,3, 5,7,10]', 'gender'))
#summarize
m7.preds
ggplot(m7.preds)
ggplot(m7.preds, aes(x=x, y=predicted, col=group))+geom_point()+geom_line()+labs(x='Ideology')
```

#### Assignment 13
> 1. Generate predicted values from model m7 for ideology, gender,and income. Plot this, finding ways to group group by income and facet by gender.Save it as m7.preds2

```{r assignment-13-solution}
m7.preds2<-ggpredict(m7, terms=c('ideology [0,3, 5,7,10]', 'income', 'gender'))
#plot
ggplot(m7.preds2, aes(x=x, y=predicted, col=group))+facet_grid(~facet)+geom_point()+geom_line()+labs(x='Ideology', col='Income')

```


