---
title: "LSIRM 2018 - Introduction to R - Tutorial 4"
author: "Simon Kiss"
date: '2018-07-12'
output: html_document
---
---
title: "LSIRM 2018 - Introduction to R - Tutorial 4"
author: "Simon Kiss"
date: '2018-07-04'
output: 
  html_document: 
    toc: true
    toc_float: true
    code_folding: hide
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=F, message=F, results='hide', fig.width=4, fig.height=3, fig.align='center')

```

# Advanced Data Management

Often, we deal with datasets that are larger than what we started with. This can pose a few problems. 

```{r load-library}
library(tidyverse)
#read in data 
ces_full<-read_sav('https://github.com/sjkiss/LSIRM_2018/raw/master/data/CES15_CPS%2BPES_Web_SSI%20Full.SAV')
```
#### Assignment 1
>1. Get a feel for the size and nature of the data set.  Store teh value labels and variable names. *do not use the names for hte same objects from the previous lesson.* 

```{r inspect}
head(ces_full)
str(ces_full)
library(labelled)

ces.vars<-var_label(ces)
ces.vals<-val_labels(ces)
```

The `questionr` package has some really useful functions to look through labelled data. 

```{r }
library(questionr)
#Sex
lookfor(ces_full, 'sex')
# 
lookfor(ces_full, 'leader')
lookfor(ces_full, 'economy')
describe(ces_full, 'economy')
```

#### Assignment 2
>1. Play around with the lookfor() command and the describe() commands 

# Advanced selecting

The `tidyverse` provides a lot of useful ways to select variables for analysis. Examine the help page for select. 

```{r select-help}
?select
```
#### ASsignment 3
```{r}
#OPtion 1
ces_full %>% 
  select(contains('emecon'))
#Optino 2
ces_full %>% 
select(p_emecon_angr:p_emecon_hope)
```

We need to deal with the missing values.  Check the values again using `lookfor()`
```{r}
describe(ces_full, 'economy')
```

It looks like anything above 4, we can recode out as missing. 


# Mutating and summarizing multiple variables

There are two variants of `mutate()` that mutate multiple variables: `mutate_all()` changes all variables, `mutate_at() changes variables using the `select()` grammar above, and `mutate_if()` changes variables if a logical condition is met. 

Inside `mutate_at()` we use two arguments `vars()` to say which variables we want to mutate and `funs` to say which function should be applied. 

```{r recode-test}
#Take the data.frame
ces_full %>% 
  #select only the emecon variables
  select(contains('emecon')) %>% 
  #We only want to mutate a subset of variables, so we select 
  mutate_all(funs(Recode(., "4:1000=NA"))) %>% 
  #Do a summary
  summary()
```


One of the really useful things about the labelled package is that you can quite quickly and easily make use of the attached value labels and turn a variable that is stored as a `double-and-labelled` variable. Normally, without labelled variables, you can use `as.factor()` to convert something to a factor.  But `labelled` variables require their own command, `as_factor()`. 

The other major advantage to this is that, after using the as_factor() command, the levels are not ordered by alphabet, they are orderd by the number they are associated with. So if 1 is `Never`, and 2 is `Always`, then the first level will be `Never` and the second will be `Always`. Using `as.factor()` from base R would reverse this because levels would be stored in alphabetical order.  

```{r}
ces_full %>% 
  #select only the emecon variables
  select(contains('emecon')) %>% 
  #We only want to mutate a subset of variables, so we select 
  mutate_all(funs(Recode(., "4:1000=NA"))) %>% 
  as_factor()
```

So, now, in four lines of code, we have: 1) dealt with missing values and 2) converted labelled variables to factors for four variables. We could do a lot of stuff with this. 

First, we can add a prefix to the recoded variables in the `funs()` argument

```{r add-previx}
ces_full %>% 
  #select only the emecon variables
  select(contains('emecon')) %>% 
  #We only want to mutate a subset of variables, so we select 
  mutate_all(funs(out=Recode(., "4:1000=NA"))) %>% 
  as_factor() %>% 
  select(contains('out')) -> econ_feeling
#Check 
econ_feeling
#Check
summary(econ_feeling)
```
This is convenient because now we can join the two data sets and add the recoded columns to the original data frame. 

Or, and this would be even cooler, we can source the previous tutorial and join the two. 

```{r}

```




# Multiple Linear Models

# Reporting Results (RMarkdown)